---
title: "EDS223 Homework 1"
subtitle: "Environmental (in)justice maps"
author: "Eden Gonzalez"
date: 2025-10-1
execute: 
  eval: true
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
library(sf)
library(here)
library(tmap)
library(viridisLite)
```

```{r}
# read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 
```

```{r}
# Filter to New Mexico
new_mexico <- ejscreen |> 
  filter(ST_ABBREV == "NM") 
```

```{r}
# Find the average values for all variables within New Mexico counties
new_mexico_counties <- aggregate(new_mexico, by = list(new_mexico$CNTY_NAME), FUN = mean)
```


```{r}
# Mapping the mean percentile of people of color in New Mexico counties
people_of_color <- tm_shape(new_mexico_counties) +
  tm_graticules(col = "lightgray") +         
  tm_polygons(lwd = 0.5, 
              fill = "P_PEOPCOLORPCT", 
              fill.scale = tm_scale(values = c("lightcyan1", "darkslateblue") ),
              fill.legend = tm_legend(title = "Mean Percentile of\nPeople of Color",
              position = tm_pos_out("right"), 
              frame = FALSE,
              title.size = 0.7)
             )  +
  tm_scalebar() +
  tm_compass(position = c("right", "top")) +
  tm_credits("Source: EPA") +
  tm_title("Mean Percentile of People of Color in New Mexico Counties") +
  tm_layout(frame.lwd = 0.4)
```


```{r}
air_releases <- tm_shape(new_mexico_counties) +
  tm_graticules(col = "lightgray") +
  tm_polygons(lwd = 0.5, 
              fill = "P_RSEI_AIR",
              fill.scale = tm_scale(values = c("lightcyan1", "darkslateblue")),
              fill.legend = tm_legend(title = "Mean Percentile of\nToxic Releases to Air", position = tm_pos_out("right"), 
              frame = FALSE,
              title.size = 0.7), 
              ) +
  tm_scalebar() +
  tm_compass(position = c("right", "top")) +
  tm_credits("Source: EPA") +
  tm_title("Mean Percentile of Toxic Releases to Air in New Mexico Counties") +
  tm_layout(frame.lwd = 0.4)
  

```


```{r}
air_releases
```

The data frame `ej_screen` was first filtered to keep data from New Mexico into a `new_mexico` dataframe. This new data frame was then aggregated by county to get the mean values of all variables. `new_mexico_counties` is a data frame where each row is a county, and the corresponding values are the mean values for each variable. 



`map1` maps the mean percentile of people of color in New Mexico counties. 

The map1 

- TGenerally 


```{r}
people_of_color 
```